# The data
Plotting the height versus the weight in Howell data

## Setting it up
```{r}
library(rethinking)
data("Howell1")
d<- Howell1
```

## Scaling the variables

```{r}

d2 <- d[d$age > 18,]

mH <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu ~ dnorm(178, 20),
    sigma ~ dunif(0,50)
  ), data = d2
)

precis(mH)

```

# Post modelling checks

## Extracting samples from the posterior

```{r}

post <- extract.samples(mH, n = 1e4)
str(post)

```

# Building a Linear Model

## Modeling the prior predictive distributions

```{r}

N <- 100
a <- rnorm(N, 178, 20)
b <- rlnorm(N, 0, 1)

plot(NULL, xlim=range(d2$weight), ylim=range(-100,400), xlab = 'weight', ylab = 'height')

abline(h=0, lty=2)
abline(h=272, lty=2)

for (i in 1:N)
  curve(a[i] + b[i]*(x - mean(d2$weight)),
        from = min(d2$weight),
        to = max(d2$weight), add=TRUE,
        col=col.alpha('black', 0.2))
```


## Modeling height with weight

```{r}
xbar <- mean(d2$weight)

mH.lm <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*(weight - xbar),
    a <- dnorm(178, 20),
    b ~ dlnorm(0,1),
    sigma ~ dunif(0,50)
  ), data = d2
)

precis(mH.lm)
```

## Plotting the most plausible lines implied by the above fit
```{r}
post <- extract.samples(mH.lm)
str(post)

a_map <- mean(post$a)
b_map <- mean(post$b)

plot(d2$height ~ d2$weight, col=rangi2)
for (i in 1:length(post$a))
  curve(post$a[i] + post$b[i]*(x-mean(d2$weight)), col=col.alpha("black",0.3), add=TRUE)
```

## Finding the mu interval for a given value of weight
```{r}

# extract samples from the posterior
post <- extract.samples(mH.lm)

mu_at_50 <- post$a + post$b*(50-mean(d2$weight))

dens(mu_at_50, show.HPDI = TRUE, lwd=2, col=rangi2)
```

## using the link function
We would like to generate the mu values for a specific range of values using the link function.  Link computes the mu for each sample in the posterior and for each weight in weight.seq.
```{r}

weight.seq <- seq(from=25, to=70, by=1)
mu <- link(mH.lm, data = data.frame(weight=weight.seq))
# finding mean & PI of mu
mu.mean <- apply(mu, 2 , mean)
mu.PI <- apply(mu, 2, PI)
mu.HPDI <- apply(mu, 2, HPDI, 0.89)

par(mfrow=c(1,2))
plot(height ~ weight, data=d2, col=rangi2)
# Looping and plotting the mu for each of the weight values
for (i in 1:length(weight.seq))
  points(weight.seq, mu[i,])

plot(height ~ weight, data=d2, col=rangi2)
lines(weight.seq, mu.mean)
shade(mu.HPDI, weight.seq)
```
## Simulating the heiht rather than the mean from the posterior distribution

```{r}
sim.height <- sim(mH.lm, data=data.frame(weight=weight.seq))
str(sim.height)

height.PI <- apply(sim.height, 2, PI, prob=0.89)

# raw data
plot(height ~ weight, data=d2, col=rangi2)

# MAP line
lines(weight.seq, mu.mean)

# HPDI region for mean
shade(mu.HPDI, weight.seq)

# PI region for simulated heights
shade(height.PI, weight.seq)

```

# Multivariate Bayesian Regression

##Loading the data
```{r}
data("WaffleDivorce")
dWD <- WaffleDivorce

#standardize the variables
dWD$A <- scale(dWD$MedianAgeMarriage)
dWD$D <- scale(dWD$Divorce)
dWD$M <- scale(dWD$Marriage)

```

## Model Building
Here we are trying to model the total association between age of marriage and divorce rate.  It could so happen that
divorce rate is imapct directly by age and through a *mediation* path with marriage rate.

```{r}

mA <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bA*A,
    a ~ dnorm(0,0.2),
    bA ~ dnorm(0,0.5),
    sigma ~ dexp(1)
  ), data = dWD
)

mM <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bM*M,
    a ~ dnorm(0,0.2),
    bM ~ dnorm(0,0.5),
    sigma ~ dexp(1)
  ), data = dWD
)

mAM <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bA*A + bM*M,
    a <- dnorm(0,0.2),
    bA <- dnorm(0,0.5),
    bM <- dnorm(0,0.5),
    sigma <- dexp(1)
  ), data = dWD
)

# plot(coeftab(mA, mM, mAM))

precis(mA)
precis(mM)
precis(mAM)

```

## Simulating the prior distribution
```{r}
prior <- extract.prior(mD)
A_seq <- seq(-2, 2, length.out = 100)
mu <- link(mD, post=prior, data=data.frame(A=A_seq))
str(mu)
plot(NULL, xlim=range(A_seq), ylim=c(-2,2),
     xlab='Median Marriage Age(std)',
     ylab='Divorce Rate (std)')
for (i in 1:50)
  lines(A_seq, mu[i,], col=col.alpha('black',0.3) )

```

## Posterior Residual Plots

A predictor variable residual is the average perdiction error when we use all of the other predictor variables to model a predictor of interest.

### Residual plot for Marraige Rate
Here we need to build a model, which is the regression between divorce rate and Median age (mA)

```{r}

mAr <- quap(
  alist(
    M ~ dnorm(mu, sigma),
    mu <- a + bMA * A,
    bMA ~ dnorm(0,0.5),
    a ~ dnorm(0,0.2),
    sigma ~ dexp(1)
  ), data = d
)

mu_M_A <- link(mAr)
#Average of the Divorce rate means caused by the median age in the 50 states are as follows:
mu_M_A_mean <- apply(mu_M_A, 2, mean)
mu_M_A_resid <- d$M - mu_M_A_mean
d$mu_M_A_resid <- mu_M_A_resid

# Building a residual plot
mAResDiv <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bR*mu_M_A_resid,
    a ~ dnorm(0,0.2),
    bR ~dnorm(0,0.5),
    sigma ~ dexp(1)
  ), data = d
)
res_seq <- seq(-2,2, length.out=100)
mu_R <- link(mAResDiv, data = data.frame(mu_M_A_resid=res_seq))
mu_R_mean <- apply(mu_R, 2, mean)
mu_R_PI <- apply(mu_R, 2, PI)

par(mfrow=c(1,2))

plot(M ~ A, data = d, col=rangi2,
     xlab = 'Median Age at Marraige(std)',
     ylab = 'Marraige rate (std)')
lines(d$A, mu_M_A_mean)
segments(d$A, mu_M_A_mean, d$A, d$M)


plot(d$D ~ mu_M_A_resid, col= rangi2,
     xlab = 'Marraige rate residuals',
     ylab = 'Divorce rate(std)')
abline(v=0, lty=2)
lines(res_seq, mu_R_mean)
shade(mu_R_PI,res_seq )

```




## A Counterfactual plot for multivariate regression

```{r}

# Studying counterfactual for average age and Marraige rate between -2 and 2 
M_seq <- seq(from=-2, to=3, length.out = 100)
pred_data <- data.frame(M=M_seq, A=0)

#Computre counterfactual mean divorce rate
mu_M <- link(mAM, data = pred_data)
mu_mean_M <- apply(mu_M, 2, mean)
mu_PI_M <- apply(mu_M, 2, PI)

# Simulate counterfactual divorce rates
D_sim_M <- sim(mAM, data=pred_data, n=1e4)
D_PI_M <- apply(D_sim_M, 2, PI)

# Studying counterfactual for average age and Median Age between -2 and 2 
A_seq <- seq(from=-2, to=3, length.out = 100)
pred_data <- data.frame(M=0, A=A_seq)

#Computre counterfactual mean divorce rate
mu_A <- link(mAM, data = pred_data)
mu_mean_A <- apply(mu_A, 2, mean)
mu_PI_A <- apply(mu_A, 2, PI)

# Simulate counterfactual divorce rates
D_sim_A <- sim(mAM, data=pred_data, n=1e4)
D_PI_A <- apply(D_sim_A, 2, PI)

par(mfrow=c(1,2))
# Plot the data for A=0
plot(D ~ M, data=dWD, col=rangi2, xlab='Marriage rate (standardized)',
     ylab='Divorce rate (standardized)')
mtext( "Median age marriage (std) = 0" )
lines(M_seq, mu_mean_M)
shade(mu_PI_M, M_seq)
shade(D_PI_M, M_seq)

# Plot the data for M=0
plot(D ~ M, data=dWD, col=rangi2, xlab='Median Age (standardized)',
     ylab='Divorce rate (standardized)')
mtext( "Marriage Rate (std) = 0" )
lines(M_seq, mu_mean_A)
shade(mu_PI_A, M_seq)
shade(D_PI_A, M_seq)


```

