# The data
Plotting the height versus the weight in Howell data

## Setting it up
```{r}
library(rethinking)
data("Howell1")
d<- Howell1
```

## Scaling the variables

```{r}

d2 <- d[d$age > 18,]

mH <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu ~ dnorm(178, 20),
    sigma ~ dunif(0,50)
  ), data = d2
)

precis(mH)

```

# Post modelling checks

## Extracting samples from the posterior

```{r}

post <- extract.samples(mH, n = 1e4)
str(post)

```

# Building a Linear Model

## Modeling the prior predictive distributions

```{r}

N <- 100
a <- rnorm(N, 178, 20)
b <- rlnorm(N, 0, 1)

plot(NULL, xlim=range(d2$weight), ylim=range(-100,400), xlab = 'weight', ylab = 'height')

abline(h=0, lty=2)
abline(h=272, lty=2)

for (i in 1:N)
  curve(a[i] + b[i]*(x - mean(d2$weight)),
        from = min(d2$weight),
        to = max(d2$weight), add=TRUE,
        col=col.alpha('black', 0.2))
```


## Modeling height with weight

```{r}
xbar <- mean(d2$weight)

mH.lm <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*(weight - xbar),
    a <- dnorm(178, 20),
    b ~ dlnorm(0,1),
    sigma ~ dunif(0,50)
  ), data = d2
)

precis(mH.lm)
```

## Plotting the most plausible lines implied by the above fit
```{r}
post <- extract.samples(mH.lm)
str(post)

a_map <- mean(post$a)
b_map <- mean(post$b)

plot(d2$height ~ d2$weight, col=rangi2)
for (i in 1:length(post$a))
  curve(post$a[i] + post$b[i]*(x-mean(d2$weight)), col=col.alpha("black",0.3), add=TRUE)
```

## Finding the mu interval for a given value of weight
```{r}

# extract samples from the posterior
post <- extract.samples(mH.lm)

mu_at_50 <- post$a + post$b*(50-mean(d2$weight))

dens(mu_at_50, show.HPDI = TRUE, lwd=2, col=rangi2)
```

## using the link function
We would like to generate the mu values for a specific range of values using the link function.  Link computes the mu for each sample in the posterior and for each weight in weight.seq.
```{r}

weight.seq <- seq(from=25, to=70, by=1)
mu <- link(mH.lm, data = data.frame(weight=weight.seq))
# finding mean & PI of mu
mu.mean <- apply(mu, 2 , mean)
mu.PI <- apply(mu, 2, PI)
mu.HPDI <- apply(mu, 2, HPDI, 0.89)

par(mfrow=c(1,2))
plot(height ~ weight, data=d2, col=rangi2)
# Looping and plotting the mu for each of the weight values
for (i in 1:length(weight.seq))
  points(weight.seq, mu[i,])

plot(height ~ weight, data=d2, col=rangi2)
lines(weight.seq, mu.mean)
shade(mu.HPDI, weight.seq)
```
## Simulating the heiht rather than the mean from the posterior distribution

```{r}
sim.height <- sim(mH.lm, data=data.frame(weight=weight.seq))
str(sim.height)

height.PI <- apply(sim.height, 2, PI, prob=0.89)

# raw data
plot(height ~ weight, data=d2, col=rangi2)

# MAP line
lines(weight.seq, mu.mean)

# HPDI region for mean
shade(mu.HPDI, weight.seq)

# PI region for simulated heights
shade(height.PI, weight.seq)

```

# Multivariate Bayesian Regression

##Loading the data
```{r}
data("WaffleDivorce")
dWD <- WaffleDivorce

#standardize the variables
dWD$A <- scale(dWD$MedianAgeMarriage)
dWD$D <- scale(dWD$Divorce)
dWD$M <- scale(dWD$Marriage)

```

## Model Building
```{r}

mD <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bA*A,
    a ~ dnorm(0,0.2),
    bA ~ dnorm(0,0.5),
    sigma ~ dexp(1)
  ), data = dWD
)

precis(mD)
```

## Simulating the prior distribution
```{r}
prior <- extract.prior(mD)
A_seq <- seq(-2, 2, length.out = 100)
mu <- link(mD, post=prior, data=data.frame(A=A_seq))
str(mu)
plot(NULL, xlim=range(A_seq), ylim=c(-2,2),
     xlab='Median Marriage Age(std)',
     ylab='Divorce Rate (std)')
for (i in 1:50)
  lines(A_seq, mu[i,], col=col.alpha('black',0.3) )

```

## A Counterfactual plot for multivariate regression

```{r}

# Model built
mCF <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bA*A + bM*M,
    a <- dnorm(0,0.2),
    bA <- dnorm(0,0.5),
    bM <- dnorm(0,0.5),
    sigma <- dexp(1)
  ), data = dWD
)

# Studying counterfactual for average age and Marraige rate between -2 and 2 
M_seq <- seq(from=-2, to=3, length.out = 100)
pred_data <- data.frame(M=M_seq, A=0)

#Computre counterfactual mean divorce rate
mu_M <- link(mCF, data = pred_data)
mu_mean_M <- apply(mu_M, 2, mean)
mu_PI_M <- apply(mu_M, 2, PI)

# Simulate counterfactual divorce rates
D_sim_M <- sim(mCF, data=pred_data, n=1e4)
D_PI_M <- apply(D_sim_M, 2, PI)

# Studying counterfactual for average age and Median Age between -2 and 2 
A_seq <- seq(from=-2, to=3, length.out = 100)
pred_data <- data.frame(M=0, A=A_seq)

#Computre counterfactual mean divorce rate
mu_A <- link(mCF, data = pred_data)
mu_mean_A <- apply(mu_A, 2, mean)
mu_PI_A <- apply(mu_A, 2, PI)

# Simulate counterfactual divorce rates
D_sim_A <- sim(mCF, data=pred_data, n=1e4)
D_PI_A <- apply(D_sim_A, 2, PI)

par(mfrow=c(1,2))
# Plot the data for A=0
plot(D ~ M, data=dWD, col=rangi2, xlab='Marriage rate (standardized)',
     ylab='Divorce rate (standardized)')
mtext( "Median age marriage (std) = 0" )
lines(M_seq, mu_mean_M)
shade(mu_PI_M, M_seq)
shade(D_PI_M, M_seq)

# Plot the data for M=0
plot(D ~ M, data=dWD, col=rangi2, xlab='Median Age (standardized)',
     ylab='Divorce rate (standardized)')
mtext( "Marriage Rate (std) = 0" )
lines(M_seq, mu_mean_A)
shade(mu_PI_A, M_seq)
shade(D_PI_A, M_seq)


```

